---
title: "NycFlights13"
author: "Dipta Dutta"
date: "20 November 2017"
output:
  html_document: default
  pdf_document: default
---
Exploring NYCFlights data 
This dataset is about all the flights leaving from NYC (i.e. JFK, LGA or EWR) in 2013.
```{r}
library(nycflights13)
library(tidyr)
library(dplyr)
library(rmarkdown)
library(Hmisc)
library(maps)
library(corrplot)
help(package = "nycflights13")

dim(flights)

View(flights)

glimpse(flights)

```
Flights are often delayed Performing explorartory Analysis will give some important questions to be answered

1. what was the Worst day to fly put of NYC in 2013

For that we need to find out the Total Delay = Avg Arrival Delay + Avg Departure Delay 

loading the flights data
```{r}
flights %>% 
  select(month, day, arr_delay, dep_delay) %>% 
  filter(arr_delay >= 0, dep_delay >= 0) %>%
  group_by(month, day) %>%
  summarise(avg_delay =  mean(arr_delay, na.rm = TRUE) + 
              mean(dep_delay, na.rm = TRUE)) %>%
  ungroup() %>%
  arrange(-avg_delay)
```
So we can see that September 12th is the worst day to fly out from NewYork where averagae delay was around 228 minutes

Next is we want to know is there any any particular airport where the delay happened 
```{r}
dataflights = flights %>% 
  select(origin,month,day,arr_delay,dep_delay) %>% 
  filter(arr_delay >=0 , dep_delay >=0) %>% 
  group_by(origin,month,day) %>% 
  summarise(avg_delay =  mean(arr_delay, na.rm = TRUE) + 
              mean(dep_delay, na.rm = TRUE)) %>%
  ungroup() %>%
  arrange(-avg_delay)

dataflights
```
As we can see the Flights originating from LGA at september 2 has the highest delay following by LGA and EWR at spetember 12th

Next we want to know total delay per Origin and plot it
```{r}
sumorigin = dataflights %>%
  group_by(origin) %>% 
  summarise(TotalDelay=sum(avg_delay)) %>% 
  ungroup() %>% 
  arrange(-TotalDelay)
sumorigin
```

We can see the the maximum flights originited from EWR are delayed

Next we are going to create a variable speed for the carriers 
```{r}
SpeedDta=flights %>% 
  mutate(speed = distance/air_time * 60) %>% 
  group_by(carrier) %>% 
  summarise(AvgSpeed = mean(speed,na.rm = T)) %>% 
  ungroup() %>% 
  arrange(-AvgSpeed)
SpeedDta  
```

So we can see  the fastest carrier is HA 

next we want to know is there any seasonal patterns for the delay

Creating the 'date' variable by passing the entire data object inside a 'with'
function. The 'ISOdate' function merges the 'month' & 'day' column along with
year = 2013 to make a complete date column. We save this entire column in the 
'dataflights' object.

```{r}
dataflights$date <- with(dataflights, ISOdate(year = 2013, month, day))
```

Creating a ggplot function with x-axis = Date and y-axis = Average delay
```{r}
ggplot(dataflights, aes(x = dataflights$date, y = dataflights$avg_delay, title = "Seasonality Trends")) + geom_point(aes(color = dataflights$origin))+ xlab("Date") + ylab("Average Delay (mins)") + geom_smooth(color = "Black")
```
This shows us that there is a PEAK in delays during June, July & August month and the delays generally fall down during the winter months October, November, December & January

On a day we want to know which particular time the delay occur in general ,we will need to have a plot the average delays per hour in all days across 2013. This will give us an hourly trend (if any) in the data.

```{r}
arr_delay_data <- flights %>%
  select(arr_delay, dep_delay, hour)
```


arr_delay_data[arr_delay_data$hour==24,]
Making sure that 24:00 hrs is 00:00 hrs. Replacing all instances of 2400 hrs 
as 0000hrs. This will help us in making sure we have only 24 hours
when we round off the hours to the nearest hour interval
```{r}
arr_delay_data$hour <- ifelse(arr_delay_data$hour == 2400, 0, arr_delay_data$hour)
```
to get a summarized dataset with average arrival delay for each hour.
```{r}
arr_delay <- arr_delay_data %>%
  filter(arr_delay >= 0, dep_delay >= 0) %>%
  select(hour, arr_delay, dep_delay) %>%
  group_by(hour) %>%
  summarise(avg_delay = mean(arr_delay, na.rm = TRUE) +
              mean(dep_delay, na.rm = TRUE)) %>%
  na.omit()
arr_delay
ggplot(arr_delay,aes(x = as.numeric(hour),y = avg_delay),title = "Daily Hourly") + geom_point(color = "Black") + geom_smooth() + xlab("Hour") + ylab("Average Delay")

```
we can see that the flights starting around 15 - 20 have the maximum delays 

Flights are often delayed due to weather condition so our next analysis will be on how weather affects the NYC flights

```{r}
flights = flights
weather = weather

#creating 'hours' column and making sure that hours are only from
#00:00 hrs to 23:00 hrs which is exactly as the 'weather' data set has.
#this will help us join the 2 data sets properly.
flights$hour <- ifelse(flights$hour == 24, 0, flights$hour)
#creating 'hours' column and making sure that hours are only from
#00:00 hrs to 23:00 hrs which is exactly as the 'weather' data set has.
#this will help us join the 2 data sets properly.
flights$hour <- ifelse(flights$hour == 24, 0, flights$hour)

#JOining the 'flights' and 'weather' datasets based on unique identifiers.
flights_weather <- left_join(flights, weather)

#We create a 'delays' column that is addition of all the delays in
#arrival and departure (Consider only positive delays. Reason stated in
#1st problem solution)
flights_weather$arr_delay = ifelse(flights_weather$arr_delay >= 0,
                                    flights_weather$arr_delay, 0)
flights_weather$dep_delay = ifelse(flights_weather$dep_delay >= 0,
                                    flights_weather$dep_delay, 0)
flights_weather$total_delay = flights_weather$arr_delay + flights_weather$dep_delay


#creating a data with only delay and weather columns. Removing origin, date/time
#because we want a correlation between delay and the particular weather condition.
cor_data = select(flights_weather, total_delay, temp, dewp, humid,
                   wind_dir, wind_speed, wind_gust, precip, pressure, visib)
#WE first plot a correlation Matrix using corrplot to find the variables that
#are correlated. We create a correlation matrix using 'cor' function
corrplot(cor(na.omit(cor_data)), method = "circle", type = "upper",
         tl.srt = 25, tl.col = "Black", tl.cex = 1, title = "Correlation
         between all 'weather' variables & 'delay'", mar =c(0, 0, 4, 0) + 0.1)
```

```{r}
## So from the corrplot we can see there are delay is dependent on all the weather variable
ggplot(cor_data, aes(y = humid, x = total_delay, title = "Total Delay v/s Relative Humidity")) + geom_smooth() + ylab("Relative Humidity") + xlab("Total Delay (mins)")

```

```{r}
# we can see that the total delay increases with the increase of humididty

ggplot(cor_data, aes(y = temp, x = total_delay, title = "Total Delay v/s Temperature")) + geom_smooth() + ylab("Temperature") + xlab("Total Delay (mins)")

```
```{r}
# We can see there is not much change in delay with temparature and delay 
ggplot(cor_data, aes(y = visib, x = total_delay, title = "Total Delay v/s Visibility")) + geom_smooth() + ylab("Temperature") + xlab("Total Delay (mins)")

```
we can see that as the visibilty decreases Delay also increases 

impacted by the aircraft used. Do aircrafts with certain characteristics (e.g. manufacturer) 
demonstrate better performance
We will create a variable name PI which will define the performance index with below 
Flight Perfomance = (Arival Delay + Departure Delay)/Air Time
We will create another variable called age which will be defined by the formula 
Age of plane = 2013(Flight year) - Year Manufactured

Merging the datasets 'flights' and 'planes' based on a unique idetifiers.
```{r}
flight_planes = left_join(flights, planes, by = 'tailnum')

names(flight_planes)[1] = "year"
names(flight_planes)[20] = "year_manufacture"

#Calulcating the Performance Index [pi] 
flight_planes$pi = (flight_planes$dep_delay + flight_planes$arr_delay)/ flight_planes$air_time

#We also calculate the 'age' of the aircraft as a variable
flight_planes$age = flight_planes$year - flight_planes$year_manufacture

#We group all the performance parameters for each aircraft using their tailnum
#This will give us an average performance index of that aircraft for the year 2013
aircraft = flight_planes %>%
  
  #we group by the tailnum
  group_by(tailnum) %>%
  
  #we find the mean of the Performance index
  summarise(avg_pi = mean(pi, na.rm = TRUE)) 
aircraft

#we will now add the planes dataset to our aggregated aircraft data
air_details = left_join(aircraft, planes)
air_details
#plotting to find any correlation
cor(na.omit(air_details[,unlist(lapply(air_details, is.numeric))]))
d = air_details %>%
  select(seats, avg_pi) %>%
  group_by(seats) %>%
  summarise(pi = mean(avg_pi, na.rm = TRUE))
#Plot to find the PI with the seats
ggplot(d, aes(x = seats, y = pi, title = "PI v/s seats")) + geom_point() + geom_smooth(method = 'lm') + xlab("Seats") + ylab("Performance Index (Lower is better)")
```

As we can see the the number of seats increases decreases the average performance index

Next we want to know the performance index per manufactuer

```{r}
d = air_details %>%
  select(manufacturer, avg_pi) %>%
  group_by(manufacturer) %>%
  summarise(pi = mean(avg_pi, na.rm = TRUE)) %>%
  na.omit()
ggplot(d, aes(x = factor(manufacturer) , y = pi, title = "PI v/s Manufacturers")) + 
  geom_bar(stat = "identity") + xlab("Perfomance Index (Lower is better)") +
  ylab("Manufacturers") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```


with this we can easily find which manufacturer has good perfromance index 
